---
title: "Food Price Forecasting"
author: "Sanjeev N S"
date: "2023-12-11"
output: html_document
---
<span style="color: blue;">**Importing datasets**</span>
```{r}
dataset <- read.csv("D:\\Desktop FILES\\7TH SEM\\FDA\\india.csv")

```
*The above is the main dataset which has food prices from year 2000-2019
```{r}
val <- read.csv("D:\\Desktop FILES\\7TH SEM\\FDA\\india_2019.csv")

```
*The above is the validation dataset which has food prices from year 2019-2020
<span style="color: blue;">**Loading required packages**</span>
```{r}
library(dplyr)
library(vars)
library(forecast)
library(ggplot2)
library(prophet)
library(corrplot)
library(xgboost)
```

<span style="color: blue;">**basic analysis of dataset**</span>
```{r}
head(dataset,7)
tail(dataset,5)
head(val,6)
tail(val,6)
```

```{r}
summary(dataset)
str(dataset)
```

```{r}
summary(val)
str(val)

```
<span style="color: blue;">**Missing Value analysis**</span>
```{r}
colSums(is.na(dataset))
```
```{r}
colSums(is.na(val))
```

<span style="color: blue;">**missing value imputaion with mean**</span>
```{r}
dataset <- dataset %>%
  mutate_if(is.numeric, ~ifelse(is.na(.), mean(., na.rm = TRUE), .))

colSums(is.na(dataset))
```

```{r}
val <- val %>%
  mutate_if(is.numeric, ~ifelse(is.na(.), mean(., na.rm = TRUE), .))

colSums(is.na(val))
```
<span style="color: blue;">**visualizations**</span>

```{r}
hist(dataset$Rice, main = "Rice Price Distribution", xlab = "Rice Price")

```
```{r}
ggplot(dataset, aes(y =Chickpeas )) +
  geom_boxplot(fill = "green") +
  labs(title = "Box Plot: chickpeas Prices", y = "checkpeas Price")

```

```{r}
ggplot(dataset, aes(y =Chickpeas )) +
  geom_boxplot(fill = "red") +
  labs(title = "Box Plot: Oil..mustard. Prices", y = "Oil..mustard. Price")
```

```{r}
ggplot(dataset, aes(y =Potatoes )) +
  geom_boxplot(fill = "cadetblue1") +
  labs(title = "Box Plot: Potatoes Prices", y = "Potatoes Price")
```
```{r}
ggplot(dataset, aes(y =Rice )) +
  geom_boxplot(fill = "gold1") +
  labs(title = "Box Plot: Rice Prices", y = "Rice Price")
```

```{r}
ggplot(dataset, aes(y =Sugar )) +
  geom_boxplot(fill = "darkseagreen2") +
  labs(title = "Box Plot: Sugar Prices", y = "sugar Price")
```
```{r}
ggplot(dataset, aes(y =Wheat )) +
  geom_boxplot(fill = "slategray3") +
  labs(title = "Box Plot: Wheat Prices", y = "Wheat  Price")
```

from these visualizations, we can infer that except Oil..mustard price,all other price has outliers in it.next part in EDA is to identify outliers and impute it with mean



<span style="color: blue;">**data type conversion**</span>
converting date atribute from "string" datatype to "date" datatype
```{r}
dataset$date <- as.Date(dataset$date,format="%Y-%m")
class(dataset$date)
```
```{r}
val$date <- as.Date(val$date,format="%Y-%m")
class(val$date)
```
<span style="color: blue;">**FORECASTING MODELS**</span>
```{r}
cat("\n")
```
<span style="color: blue;">**ARIMA MODEL**</span>
```{r}
# Create a time series object
rice_ts <- ts(dataset$Rice, frequency = 12, start = c(2000, 2))
arima_model <- auto.arima(rice_ts)
forecast_values <- forecast(arima_model, h = 12)
print(forecast_values)
```


```{r}
point_forecasts <- forecast_values$mean
point_forecasts
error_arima <- abs(sum(point_forecasts-val$Rice)/12)
error_arima
```
<span style="color: blue;">**Exponential Smoothing (ETS)**</span>
```{r}
ets_model <- ets(rice_ts)
forecast_values_ets <- forecast(ets_model, h = 12)
print(forecast_values_ets)
```

```{r}
point_forecasts <- forecast_values_ets$mean
point_forecasts
error_ETS <- abs(sum(point_forecasts-val$Rice)/12)
error_ETS
```
<span style="color: blue;">**SARIMA (Seasonal ARIMA)**</span>
```{r}
sarima_model <- arima(rice_ts, order = c(1, 1, 1), seasonal = list(order = c(2, 1, 1)), method = "CSS")
forecast_values_sarima <- forecast(sarima_model, h = 12)
print(forecast_values_sarima)
```

```{r}
point_forecasts <- forecast_values_sarima$mean
error_SARIMA <- abs(sum(point_forecasts-val$Rice)/12)
error_SARIMA
```
<span style="color: blue;">**VECTOR AUTOREGRESSION MODEL[VAR]**</span>
```{r}
multivariate_data <- dataset[, c("Rice", "Wheat")]

# Convert the data to time series format
ts_data <- ts(multivariate_data, frequency = 12, start = c(2000, 2))

# Fit VAR model
var_model <- VAR(ts_data, p = 1)  # Specify the lag order (p) based on your data


forecast_horizon <- 12 
forecast_values_var <- predict(var_model, n.ahead = forecast_horizon)[]

```


```{r}
# Print the point forecast values
print(forecast_values_var$fcst$Rice[,1])
```
```{r}
error_VAR <- abs(sum(forecast_values_var$fcst$Rice[,1]-val$Rice)/12)
error_VAR
```

<span style="color: blue;">**validation**</span>
<span style="color: blue;">comparision of forecasting models</span>
```{r}
cat("ARIMA ERROR: ", error_arima, "\n")
cat("Exponential Smoothing ERROR: ", error_ETS, "\n")
cat("SARIMA ERROR: ", error_SARIMA, "\n")
cat("Vector Autoregression (VAR): ", error_VAR, "\n")
```
```{r}
best_model <- ifelse(abs(error_arima) <= abs(error_SARIMA )& abs(error_arima) <= abs(error_ETS) & abs(error_arima) <= abs(error_VAR), "ARIMA",
                   ifelse(abs(error_ETS) <= abs(error_arima) & abs(error_ETS) <= abs(error_SARIMA) & abs(error_ETS) <= abs(error_VAR), "Exponential Smoothing",
                   ifelse(abs(error_VAR) <= abs(error_SARIMA) & abs(error_VAR) <= abs(error_ETS), "Vector Autoregression (VAR)", "SARIMA(Seasonal ARIMA)")))

print(best_model)

```
```{r}

```
```{r}
models <- c("ARIMA", "Exponential Smoothing", "SARIMA","Vector Autoregression (VAR)")
error_values <- c(error_arima,error_ETS,error_SARIMA,error_VAR)

# Create a data frame
error_data <- data.frame(Model = models, ERROR = error_values)

ggplot(error_data, aes(x = Model, y = ERROR)) +
  geom_bar(stat = "identity", fill = "yellow2") +
  labs(title = "ERROR for Different forecasting Models",
       x = "Model",
       y = "ERROR") +
  theme_minimal()
```

<span style="color: black;/n">we can observe visually that SARIMA model has less error than other models,so SARIMA model is a best model to forecast the prices</span>

```{r}
combined_forecast <- cbind(
  arima = forecast_values$mean,
  ets = forecast_values_ets$mean,
  sarima = forecast_values_sarima$mean
  #var = forecast_values_var$fcst$Rice[, 1]
)
```
```{r}
# Train gradient boosting model
xgb_model <- xgboost(data = as.matrix(combined_forecast), label = as.numeric(head(rice_ts, nrow(combined_forecast))), nrounds = 100, objective = "reg:squarederror")

# Make final forecast
final_forecast <- predict(xgb_model, as.matrix(combined_forecast))

# Print final forecast
print(final_forecast)

```


```{r}
point_forecasts <- final_forecast
error_ensemble <- abs(sum(point_forecasts-val$Rice)/12)
abs(error_ensemble)
```
```{r}
models <- c("ARIMA", "Exponential Smoothing", "SARIMA","Ensemble model")
error_values <- c(error_arima,error_ETS,error_SARIMA,error_ensemble)

# Create a data frame
error_data <- data.frame(Model = models, ERROR = error_values)

ggplot(error_data, aes(x = Model, y = ERROR)) +
  geom_bar(stat = "identity", fill = "yellow2") +
  labs(title = "ERROR for Different forecasting Models",
       x = "Model",
       y = "ERROR") +
  theme_minimal()
```


